Ecological Limits to Forecasting: Estimating Uncertainty in Critical Transitions with Deep Learning

### 2022-05-16 manuscript/manuscript.Rmd
:pen:

```{r }
#| fig.cap = "Forecasts of the saddle-node bifurcation under each model, compared to 15 realizations of the true model.
#| Training data proceeds the bifurcation, making accurate prediction without knowledge of the underlying model very difficult."

@@

@@

In this paper, we carry out an initial exploration on how deep learning methods can perform on predicting transition events.
We compare the ability of several cutting edge machine learning approaches against statistical and process-based models, and show that deep learning methods are generally able to strike a middle ground between what we consider as acceptable and ideal case forecasting methods, ARIMA and MCMC-based parameter estimation respectively.
When machine learning models are able to observe transition phenomena, as in the stochastic case, they generally outperformed MCMC-based forecasting with respect to CRPS and log probability score but under-performed MCMC when there were no transition events in the training sets as in the Hopf and Saddle-node examples.
Yet, there are many nuances to these assessments as examining summary statistics, CRPS and log probability scores obscure finer detailed components of the forecasts, like the pronounced bimodality of ML forecasts for the stochastic transition case.

>>>>

```{r }
#| fig.cap = "Forecasts of the saddle-node bifurcation under each model, compared to 15 realizations of the true model. 
#| Training data proceeds the bifurcation, making accurate prediction without knowledge of the underlying model very difficult. "

@@

Examining patterns in the scores over time (Fig 6-7) provides a more nuanced understanding of the forecast dynamics than aggregate scores alone (Fig 2).
In the Hopf bifurcation, CRPS scores get worse over time across all methods, including the MCMC forecasts.
In the saddle node bifurcation and stochastic transition, the same pattern holds somewhat more dramatically for non-MCMC forecast,
while MCMC scores are worst in mid-range.
Comparing CRPS scores to logs score also emphasizes the relative role of uncertainty: for instance, the MCMC scores for the Hopf bifurcation get steadily worse under CRPS but not under logs score. 
A relatively sharp transition can be seen under both MCMC scores on the Hopf bifurcation once the magnitude of the oscillations exceeds the variance created by mere stochasticity: the MCMC model no longer over-estimates the spread of the data, while the ML models now underestimate that variation. 
CRPS scores for stochastic transitions exhibit a distinct two-branch pattern, with scores for a given replicate being either very high (poor skill) or very low, reflecting whether the individual 'true' replicate matches the mean state predicted by the forecast or the other state. 
Logs skill score may be a better measure in this context, where correctly capturing the uncertainty in the forecast means that this bi-modal structure in scores can be avoided entirely, e.g. by the MCMC predictions. 
The forecast-skill-over time plots illustrate diferent reasons for the bi-modal distribution in skill seen for the saddle-node and stochastic transition scenarios in Figure 2 respectively: in the case of the saddle node, the two modes are distinguished by time-horizon: short term forecasts are relatively accurate, longer term forecasts (i.e. after the catastrophic transition) are poor.
In the case of the stochastic transition model, the two modes are not structured by horizon but by replicate, with some replicates having transitioned and others still in the original state.

@@

In this paper, we carry out an initial exploration on how deep learning methods can perform on predicting transition events.
We compare the ability of several cutting edge machine learning approaches against statistical and process-based models, and show that deep learning methods are generally able to strike a middle ground between what we consider as acceptable and ideal case forecasting methods, ARIMA and MCMC-based parameter estimation respectively.
All though most ML-based forecasting applications focus on point predictions, we have emphasized examples which can provide a meaningful estimate of uncertainty.
When machine learning models are able to observe transition phenomena, as in the stochastic case, they performed comparably to MCMC-based forecasting with respect to CRPS and log probability score but under-performed MCMC when there were no transition events in the training sets as in the Hopf and Saddle-node examples.
An ensemble forecast combining the predictions of all four ML methods generally scores as well or better than any one of the ML methods alone.
Yet, there are many nuances to these assessments as examining summary statistics, CRPS and log probability scores obscure finer detailed components of the forecasts.
For instance, forecast skill varies with the length of the forecast horizon, and not in a monotonic fashion. 
This is the result of multiple factors: for some dynamics, such as those involving tipping points, the long term behavior can be easier to predict then transient transitions. 
Both predicted uncertainty and forecast skill can be better on longer horizons than on shorter ones, as in the MCMC predictions of tipping point dynamics.
It is also important to remember that probabilistic forecast skill scores do not only measure how close observations are to expectation, but also reflect the predicted uncertainty: therefore, over-confidence about predictive accuracy can result in worse scores than less accurate on average but correctly reflect a greater degree of uncertainty. 
The ability to better reflect uncertainty rather than better average predictions explains much of the performance of the ensemble model.

The success of these ML models on the stochastic transition case is particularly notable.
Using only a single previous replicate

Overall, ML models and the more traditional ARIMA model fail to predict the qualitative shift in dynamic behavior that occurs in these critical transitions.
This is not surprising, as the training data provide no prior example of such behavior (e.g. growing oscillations or a sudden shift).
Nevertheless, this should be an important reminder of a central difficulty in ecological forecasting.
Note that in such scenarios, near-term forecasts [@dietze_iterative_2018] may be very accurate right up to the transition event before becoming widely wrong. 
Nor can the possibility of such non-linear behavior be easily dismissed in ecological models -- the examples considered here have been bedrock of ecological modeling and management practices for over half a century [@Folke2004], and if anything are only too simple, representing a small slice of possible dynamical behavior of more complicated models.

Identifiability
